Co-moving stars in \textsl{Gaia DR1}

### 2016-11-02 paper/ms.tex
rewrote data; added new figures

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\mat}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bs{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}
\newcommand{\eye}{\mathbb{I}}

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (\tgas), released as a part of Data Release 1 (\dr) of the Gaia mission
\citep{2016arXiv160904172G,2016arXiv160904303L}.
\tgas\ contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog \citep{2000A&A...355L..27H} with median
astrometric precision comparable to that of the \project{Hipparcos} catalog
\citep[$\approx 0.3~{\rm mas}$;][]{2007ASSL..350.....V}. In terms of parallax
signal-to-noise ($\snr = \varpi/\sigma_\varpi$), the \tgas\ catalog contains
42385 high-precision stars with $\snr > 32$: here we visualize quantities
computed from this high-precision sample to emphasize the quality of the data.
In the sections that follow, we construct a statistical model that propagates
the non-trivial uncertainties in the data to our beliefs about the likelihood
that a given pair of stars is co-moving.

For each star with $\snr > 32$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 29870 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \vec{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows the tangential velocity difference against the
physical separation plotted for all pairs that pass the above signal-to-noise
cut. Pairs with small physical separation also tend to have small velocity
differences. Intriguingly, there appear to be an abundance of pairs with small
velocity separation at separations $|\delta\vec{x}| \gtrsim 1~{\rm pc}$. Of
course, this figure really only highlights close pairs with small angular
separations:
At large angular separations, the tangential velocity differences will be large
because of projection effects.
To select a more representative population of co-moving stars, we must instead
model the 3-space velocities of the stars given observations of just two
components.

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/dv-sep.pdf}
  \end{center}
  \caption{%
    Difference in tangential velocity vs. physical separation for pairs of stars
    selected to be nearest neighbors in separation. Dashed line shows zero velocity
    difference. Note the points with small velocity difference: these are likely
    widely-separated binary stars.
    \label{fig:dv-sep}}
\end{figure*}

@@

The abundance of pairs of stars with small velocity difference in the
high-precision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\mat{C}$.
    \todo{SMOH: mention GAIA paper that says Gaussian is good approximation?}
  \item We assume that the 3-space velocities of a given pair of stars
    $(\vec{v}_i, \vec{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\vec{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\vec{v})$.
\end{itemize}

@@

\section{Results 1: A catalog of candidate wide binaries}
In this section, we examine the distribution of likelihood ratios
$\ln \mathcal{L}_1 /\mathcal{L}_2$, and come up with a reasonable cut
for this quantities to select co-moving pairs from candidate pairs list.
We present the statistics of the resulting co-moving pairs.
We also match the pairs with known open clusters in the Milky Way to
provide information on cluster membership.

@@

@@

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/hist_Nneighbors.pdf}
  \end{center}
  \caption{%
    Histogram of the number of co-moving neighbors to a star for varying
    degrees of contamination due to random pairs.
    \todo{SMOH: What is the optimal bins for this histogram to separate
    small number cases e.g., binary case (N=1) appropriately?}
    \label{fig:hist_Nneighbors}}
\end{figure*}

@@

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/glon_d_pairlines.pdf}
    \includegraphics[width=\textwidth]{figures/glon_d_pairlines_mutexc.pdf}
  \end{center}
  \caption{%
    Top: Distribution of co-moving pairs in galactic longitude and
    distance. Each co-moving pair is connected by a line.
    Red circles mark the locations of 135 known Milky Way open clusters within
    $d<600$~pc from \citealt{Kharchenko:2016aa} while
    stars in known OB associations \citep{de-Zeeuw:1999aa} are colored
    by which association they belong.
    We see that some of the aggregates of co-moving pairs are clearly
    associated with known open clusters, and OB associations.
    Bottom: Same as the previous plot but only for mutually exclusive pairs.
    The mutually exclusive pairs no longer traces any known co-moving structures
    strongly.
    \label{fig:glon_d_pairlines}}
\end{figure*}

@@

\section{Results 2: Star clusters and cluster remnants}

>>>>

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\mat}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bs{#1}}
\newcommand{\kms}{\ensuremath{\rm km~s^{-1}}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}
\newcommand{\eye}{\mathbb{I}}

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (\tgas), released as a part of Data Release 1 (\dr) of the Gaia mission
\citep{2016arXiv160904172G,2016arXiv160904303L}.
\tgas\ contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog \citep{2000A&A...355L..27H} with median
astrometric precision comparable to that of the \project{Hipparcos} catalog
\citep[$\approx 0.3~{\rm mas}$;][]{2007ASSL..350.....V}. In terms of parallax
signal-to-noise ($\snr = \varpi/\sigma_\varpi$), the \tgas\ catalog contains
42385 high-precision stars with $\snr > 32$.

We construct a sample of candidate pair list as follows.
We first apply a global parallax S/N cut of 8 to \tgas, which leaves 619,618 stars.
Then, for each star, we search a potential co-moving partner star with a
difference in tangential velocity $\Delta v_t < 10$~\kms\ within 10 pc
using RA, DEC, a point-estimate of distance, and the proper motions.
Throughout this \documentname\, we use the following estimator for
the point-estimate of distance applying a correction for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
where $\varpi$ is the parallax in mas.
The difference in tangential velocity between two stars is, then,
\begin{equation}
  |\Delta \vec{v}_{\rm t}| = |d_1 \vec\mu_1 - d_2 \vec\mu_2|
\end{equation}
where $\vec\mu = (\mu_\alpha, \mu_\delta)$.

Figure~\ref{fig:dv-sep} shows $|\Delta v_t|$ against the physical separation
for the resulting 271,232 unique candidate pairs.
A few key observations can be made:

\begin{itemize}
  \item At small separation ($<1$~pc), there is a population of pairs with
  very small tangential velocity difference ($<2$~km/s). Given that these pairs
  are very close in both positions and proper motions,
  it is highly probable that they are actually co-moving e.g., as wide binaries.

  \item  However, it is not only (wide) binaries which are gravitationally
  bound that are expected to appear as co-moving.
  OB associations, moving groups, and open clusters are also of consideration.
  Since the basic unit of our choice is a pair of stars, we can expect that these astrophysical objects may be detected as networks of pairs.
  As the pair separation increases, the nature of co-moving pairs
  will change from binaries to those related to these larger objects,
  which will tend to subtend a larger angle in sky.
  Under the assumption of having the same 3D velocity,
  two stars of a co-moving pair can essentially be thought of as projections
  of this velocity onto sphere at different viewing angles.
  The larger the difference in viewing angles is, the larger the difference in tangential 
  velocities will be.
  Due to this projection effect, a population of genuine co-moving pairs
  will extend to larger $\Delta v_t$ at larger separation.
  This indeed can be seen in Figure~\ref{fig:dv-sep} as an over-density
  in lower right corner that gets thinner as $\Delta v_t$ increases.

  \item Finally, there is a population of “random” pairs of field stars
  which are not actually co-moving, but still has $\Delta v_t < 10$~\kms\
  by chance.
  As $\Delta v_t$ increases, this population will dominate.
  However, it is clear from this Figure that there is an overlap between
  genuine co-moving pairs and “random” pairs.
\end{itemize}

In the following section, we construct a statistical model that propagates
the non-trivial uncertainties in the data to our beliefs about the likelihood
that a given pair of stars is co-moving.

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/sep_dvtan.pdf}
  \end{center}
  \caption{%
    Difference in tangential velocity vs. physical separation for candidate
    pairs of stars with separation $< 10$~pc and $|\Delta v_t| < 10$~\kms.
    % Note the points with small velocity difference: these are likely
    % widely-separated binary stars.
    \label{fig:dv-sep}}
\end{figure*}

@@

The abundance of pairs of stars with small velocity difference in
\figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at a range
of separations.
Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\mat{C}$.
    \todo{SMOH: mention GAIA paper that says Gaussian is good approximation?}
  \item We assume that the 3-space velocities of a given pair of stars
    $(\vec{v}_i, \vec{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\vec{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\vec{v})$.
\end{itemize}

@@

\section{Results}

This section is divided into three parts. First, we discuss and justify a cut of
the likelihood ratio to select candidate co-moving pairs.
Second, we present the statistics and properties of co-moving pairs, and discuss its nature.
Finally, we describe our catalog.

\subsection{Selecting candidate co-moving pairs}

In this section, we examine the distribution of likelihood ratios
$\ln \mathcal{L}_1 /\mathcal{L}_2$, and come up with a reasonable cut
for this quantities to select co-moving pairs from candidate pairs list.

With these remarks in mind, we discuss the likelihood ratio distribution.

@@

\subsection{Statistics, properties, and nature of co-moving pairs}

We present the statistics of the resulting co-moving pairs.
We also match the pairs with known open clusters in the Milky Way to
provide information on cluster membership.

@@

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/dist_networksize.pdf}
  \end{center}
  \caption{%
    Histogram of co-moving network sizes.
    \label{fig:hist_Nneighbors}}
\end{figure*}

@@

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/glon_d_pie.pdf}
  \end{center}
  \caption{%
    Panoramic view of co-moving pairs of stars. The angle is the Galactic
    longitude, and the distance is in pc.
    Pair networks of size 2 and 3 are connected by gray and blue lines.
    Pair networks of size 4 and larger are plotted with a unique color.
    Black circles mark the positions of known Milky Way star clusters
    from \citet{Kharchenko:2016aa}.
    \label{fig:glon_d_pairlines}}
\end{figure*}

% \begin{figure*}[p]
%   \begin{center}
%     \includegraphics[width=\textwidth]{figures/glon_d_pairlines.pdf}
%     \includegraphics[width=\textwidth]{figures/glon_d_pairlines_mutexc.pdf}
%   \end{center}
%   \caption{%
%     Top: Distribution of co-moving pairs in galactic longitude and
%     distance. Each co-moving pair is connected by a line.
%     Red circles mark the locations of 135 known Milky Way open clusters within
%     $d<600$~pc from \citealt{Kharchenko:2016aa} while
%     stars in known OB associations \citep{de-Zeeuw:1999aa} are colored
%     by which association they belong.
%     We see that some of the aggregates of co-moving pairs are clearly
%     associated with known open clusters, and OB associations.
%     Bottom: Same as the previous plot but only for mutually exclusive pairs.
%     The mutually exclusive pairs no longer traces any known co-moving structures
%     strongly.
%     \label{fig:glon_d_pie}}
% \end{figure*}

@@

\subsection{Catalog of candidate co-moving pairs}
